---
title: 'CTA "L" Ridership Figures Analyzed'
output:
  md_document:
    pandoc_args: ["-s", "-t", "markdown"]
layout: post
---

## Introduction

The [Chicago Transit Authority](http://www.transitchicago.com/) (CTA) provides a lot of open data about its train system (also known as the "L") and its bus system through its [data portal](http://www.transitchicago.com/data/).

We are going to look at the train ridership data.

```{r load-packages, message=FALSE, echo=FALSE}
#The analysis that follows was performed using R with the following libraries:
library(knitr)
library(dplyr)
library(ggplot2)
library(xtable)
```

```{r read-date, echo=FALSE}
cta <- read.csv('CTA_-_Ridership_-__L__Station_Entries_-_Daily_Totals.csv')
cta_stations <- read.csv('CTA_-_System_Information_-_List_of__L__Stops.csv')
```

```{r set-fig-path, echo=FALSE}
opts_knit$set(base.dir = getwd(), base.url = '/')
opts_chunk$set(fig.path = 'assets/cta/')
```

```{r add-line, echo=FALSE}
map_line <- function(booleans, name) {
  to_string <- function(x) { return(ifelse(x == "true", paste(name, ""), "")) }
  return(sapply(booleans, to_string))
}

extract_lines <- function(dataset) {
  red <- map_line(dataset$RED, "Red")
  blue <- map_line(dataset$BLUE, "Blue")
  green <- map_line(dataset$G, "Green")
  brown <- map_line(dataset$BRN, "Brown")
  purple <- map_line(dataset$P, "Purple")
  yellow <- map_line(dataset$Y, "Yellow")
  pink <- map_line(dataset$Pnk, "Pink")
  orange <- map_line(dataset$O, "Orange")
  return(paste0(red, blue, green, brown, purple, yellow, pink, orange))
}
```

```{r process-data, echo=FALSE}
cta_stations <- cta_stations %>% group_by(MAP_ID) %>% do(head(., n = 1)) %>% ungroup()
cta_stations <- cta_stations %>% mutate(line = extract_lines(.))
cta <- inner_join(cta, cta_stations, by = c("station_id" = "MAP_ID"))
```

## Dataset

The variables we are going to use are: 
<ul>
  <li>Station</li>
  <li>Date</li>
  <li>Day Type</li>
  <li>Rides</li>
  <li>Product</li>
</ul>

As a preprocessing step, data for the current year (2017) was removed because we are comparing full years.

```{r, echo=FALSE}
add_year <- function(dataset) {
  return(dataset %>%
    mutate(year = strtoi(substr(dataset$date, 7, 11))))
}

cta <- add_year(cta)
cta_to_2016 <- cta %>% filter(year != 2017)

group_by_year <- function(dataset) {
  return(dataset %>%
    group_by(year) %>%
    summarise(rides = sum(rides)))
}
```

## Ridership by Year

```{r ridership-by-year, echo=FALSE}
by_year <- cta %>% 
  filter(year > 2000) %>%
  group_by(year) %>%
  summarize(rides = sum(rides))

rides_for_year <- function(queryYear) {
  return(subset(by_year, year == queryYear, select = rides)[[1]])
}

three_year_avg <- function(queryYear) {
  return(mean(c(rides_for_year(queryYear), rides_for_year(queryYear - 1), rides_for_year(queryYear - 2))))
}
```

The CTA has seen increasing ridership in the past few years. Last year (2016) was unusually the lowest since 2011, but it could be just a temporary deviation from the upward trend. In the last 10 years, ridership has increased by `r ((three_year_avg(2016) / three_year_avg(2006)) - 1) * 100 `%.

```{r ridership-by-year-graph, echo=FALSE}
ggplot(by_year, aes(year, rides)) +
  geom_bar(stat="identity") +
  ggtitle(label = "Total Rides by Year", subtitle = NULL) +
  labs(x = "Year", y = "Rides") +
  scale_x_continuous(breaks = pretty(by_year$year, n = 10))
```

## Busiest Days

Here are the all time busiest dates on the CTA. 

<div markdown="0">
```{r, echo=FALSE}
by_date <- cta %>% 
  group_by(date) %>%
  summarize(rides = sum(rides))

sorted <- by_date %>% 
  arrange(desc(rides))
kable(sorted[1:10,], 
      caption = "Top 10 Days",
      format = 'html',
      table.attr = "class=\"table\"")
```
</div>

## Busiest Station Days

<div markdown="0">
```{r, echo=FALSE}

sorted <- cta %>% 
  arrange(desc(rides)) %>%
  select(date, stationname, line, rides)
kable(sorted[1:10,], 
      caption = "Top 10 Daily Rides",
      format = 'html',
      table.attr = "class=\"table\"")
```
</div>

## Busiest Stations

<div>
```{r, echo=FALSE}
by_station <- cta %>% 
  group_by(stationname, line) %>%
  summarize(rides = sum(rides))

sorted <- by_station %>% 
  arrange(desc(rides))
kable(sorted[1:10,], 
      caption = "Top 10 Stations",
      format = 'html',
      table.attr = "class=\"table\"")
```
</div>

## Day Types
```{r, echo=FALSE}
#TODO divide the weekdays by five to get an average
data <- cta %>% 
  group_by(daytype, year) %>%
  summarize(rides = sum(rides))

ggplot(data, aes(year, rides)) +
  geom_bar(aes(fill = daytype), position = "dodge", stat="identity") +
  ggtitle(label = "Daytype Year", subtitle = NULL) +
  labs(x = "Year", y = "Rides") +
  scale_x_continuous(breaks = pretty(data$year, n = length(unique(data$year))))
```

