---
title: "Trends in the JODI Dataset"
output:
  html_document:
    toc: true
    theme: united
---

# Dataset

In this write up, we are going to explore data provided by the Joint Organisations Data Initiative (JODI), which is an international organization that provides open data about oil and natural gas. The first data gathering exercise was launched in 2001 and many countries participated, adding up to an estimated coverage of about 90% world supply. 

The dataset we are going to use is the World Primary dataset, which can be downloaded from the [Data Downloads](https://www.jodidata.org/oil/database/data-downloads.aspx) page on the JODI website.
The World Primary dataset contains data about Crude Oil, Natural Gas Liquids (NGL), and other primary sources (Other). The World Secondary dataset contains data about secondary products such as Gas/diesel oil, Fuel oil, etc.

Disclaimer: I am not an energy expert. Most of what I know comes from reading [The Quest](https://www.amazon.com/Quest-Energy-Security-Remaking-Modern/dp/0143121944) by Daniel Yergin, a book that I would recommend to anybody who wants to learn about the history of different energy sources and the geopolitical implications of energy.

```{r load-jodi, echo=FALSE}
jodi <- read.csv('world_Primary_CSV.csv')
```

The analysis that follows was performed using R with the following libraries:
```{r load-packages, message=FALSE}
library(knitr)
library(dplyr)
library(ggplot2)
```

The variables we are going to use are: country, year, and quantity, product, flow, and unit. As a preprocessing step, data about the current year (2017) was removed because it is not complete. Also, we are starting starting our graphs from 2007 so that they can fit nicely on the screen.

```{r, echo=FALSE}
add_year <- function(dataset) {
  return(dataset %>%
    mutate(year = strtoi(substr(dataset$date, 4, 8))) %>%
    filter(year != 2017))
}

jodi <- add_year(jodi)
jodi <- filter(jodi, year > 2006)

group_by_year <- function(dataset) {
  return(dataset %>%
    group_by(year) %>%
    summarise(quantity = sum(quantity)))
} 
```

## Products

One of the data fields is Product, which is divided in Crude Oil, Natural Gas Liquids (not to be confused with Natural Gas), Other Crude, and Total Crude, which acts as the sum of the prior three. For some reason Total Crude is missing from some years. So we will be using Crude Oil, which is the major component anyway.

```{r, echo=FALSE}
jodi_us <- filter(jodi, country == 'USA', unit == 'KBBL', flow == 'PRODREFOUT', !is.na(quantity))

jodi_us_prod_by_year <- jodi_us %>%
  group_by(year, product) %>%
  summarise(quantity = sum(quantity)) %>%
  mutate(quantity = quantity / 1000)

ggplot(jodi_us_prod_by_year, aes(year, quantity)) +   
  geom_bar(aes(fill = product), position = "dodge", stat="identity") +
  ggtitle(label = "US All Oil Products", subtitle = NULL) +
  labs(x = "Year", y = 'Quantity (Millions of Barrels)') +
  scale_x_continuous(breaks = pretty(jodi_us_prod_by_year$year, n = 10))
```

## World Production

In this section we are going to first look at the world prodution and then drill down at the country level to see which ones are the top producers. We will also see what were the biggest changes in recent times.


The world 

```{r, echo=FALSE}
jodi_prod_and_use <- filter(jodi, product == 'CRUDEOIL', unit == 'KBBL', !is.na(quantity))
by_year <- jodi_prod_and_use %>%
  mutate(quantity = quantity / 1000) %>%
  group_by(year, product) %>%
  summarise(quantity = sum(quantity))

ggplot(by_year, aes(year, quantity)) +   
  geom_line(stat="identity") +
  ggtitle(label = "Production and Direct Use", subtitle = NULL) +
  labs(x = "Year", y = 'Quantity (Millions of Barrels)') +
  scale_x_continuous(breaks = pretty(by_year$year, n = 10))
```

### Top Producers

Three countries are definitely the top producers: Saudi Arabia, Russia, and the United States. Iraq, China, and Iran complete the top 5 across the last three years, although their output is around a third of Saudi Arabia.

Iraq has come up a lot. Venezuela has come down. Expand on this.

```{r, echo=FALSE}
country_yearly_stats <- function(dataset, yearToFilter) {
  return(dataset %>%
    filter(year == yearToFilter) %>%
    group_by(country))
}

jodi_prod <- filter(jodi, flow == 'PRODREFOUT', product == 'CRUDEOIL', unit == 'KBBL')
jodi_prod <- mutate(jodi_prod, quantity = quantity / 1000)

table2016 <- country_yearly_stats(jodi_prod, 2016) %>%
  summarise(quantity2016 = sum(quantity)) %>%
  mutate(quantity2016 = round(quantity2016)) %>%
  arrange(desc(quantity2016))

table2015 <- country_yearly_stats(jodi_prod, 2015) %>%
  summarise(quantity2015 = sum(quantity)) %>%
  mutate(quantity2015 = round(quantity2015)) %>%
  arrange(desc(quantity2015))

table2014 <- country_yearly_stats(jodi_prod, 2014) %>%
  summarise(quantity2014 = sum(quantity)) %>%
  mutate(quantity2014 = round(quantity2014)) %>%
  arrange(desc(quantity2014))

table2013 <- country_yearly_stats(jodi_prod, 2013) %>%
  summarise(quantity2013 = sum(quantity)) %>%
  mutate(quantity2013 = round(quantity2013)) %>%
  arrange(desc(quantity2013))

with_2015 <- left_join(table2016, table2015, by = "country")
with_2014 <- left_join(with_2015, table2014, by = "country")
with_2013 <- left_join(with_2014, table2013, by = "country")

kable(with_2013[1:15,], caption = 'Top 15 Crude Oil Producers (Millions of Barrels)', col.names = c("Country", "2016", "2015", "2014", "2013"))
```

If we only look at the three major producers, we can see that the United States has recovered a lot of ground in the last few years:

```{r, echo=FALSE}
jodi_prod <- filter(jodi, flow == 'PRODREFOUT', product == 'CRUDEOIL', unit == 'KBBL', 
                    country == 'USA' | country == 'RUSSIA' | country == 'SARABIA')
quantity_by_year_and_country <- jodi_prod %>%
  mutate(quantity = quantity / 1000) %>%
  group_by(year, country) %>%
  summarise(quantity = sum(quantity))

ggplot(quantity_by_year_and_country, aes(year, quantity)) +   
  geom_bar(aes(fill = country), position = "dodge", stat="identity") +
  ggtitle(label = "Crude Oil Top 3 Producers", subtitle = NULL) +
  labs(x = "Year", y = 'Quantity (Millions of Barrels)') +
  scale_x_continuous(breaks = pretty(quantity_by_year_and_country$year, n = 10))
```

# Individual Countries

```{r, echo=FALSE}
country_yearly_graph <- function(queryCountry) {
  jodi_prod <- filter(jodi, country == queryCountry, 
                       flow == 'PRODREFOUT' | flow == 'TOTEXPSB' | flow == 'TOTIMPSB', 
                       product == 'CRUDEOIL', unit == 'KBBL')

jodi_us_prod_by_year <- jodi_prod %>%
  group_by(year, flow) %>%
  summarise(quantity = sum(quantity)) %>%
  mutate(quantity = quantity / 1000)

ggplot(jodi_us_prod_by_year, aes(year, quantity)) +   
  geom_bar(aes(fill = flow), position = "dodge", stat="identity") +
  ggtitle(label = "", subtitle = NULL) +
  labs(x = "Year", y = 'Quantity (Millions of Barrels)') +
  scale_x_continuous(breaks = pretty(jodi_us_prod_by_year$year, n = 10)) +
  scale_fill_discrete(
    name = "Flow",
    breaks = c("PRODREFOUT", "TOTEXPSB", "TOTIMPSB"),
    labels = c("Production", "Exports", "Imports"))
} 
```

## Production, Imports, and Exports

### Saudi Arabia

Saudi Arabia is the world top producer.

```{r, echo=FALSE}
country_yearly_graph('SARABIA')
```

### Russia

```{r, echo=FALSE}
country_yearly_graph('RUSSIA')
```

### United States

The United States has seen a boom in oil production after 2010, which is commonly attributed to the technological advances in horizontal drilling and hydraulic fracturing. The graph also shows declining imports and rising exports, although they are still very small when compared to production and imports. 
 
```{r, echo=FALSE}
country_yearly_graph('USA')
```

### Iraq

```{r, echo=FALSE}
country_yearly_graph('IRAQ')
```

### Brazil

```{r, echo=FALSE}
country_yearly_graph('BRAZIL')
```

### Venezuela

```{r, echo=FALSE}
country_yearly_graph('VENEZOPEC')
```
