---
title: "Electricity by Source in the United States"
output:
  md_document:
    pandoc_args: ["-s", "-t", "markdown"]
layout: post
---

## Introduction

This post is about electricity generation in the United States. We will look at the sources used to produce electricity. The data is provided by the U.S. Energy Information Administration ([EIA](https://www.eia.gov/)) which publishes reports and statistics related to the energy sector in the United States.

The specific dataset used in this post was extracted from the page  [Net Generation by Energy Source](https://www.eia.gov/electricity/monthly/epm_table_grapher.php?t=epmt_1_01) on the EIA's website. The data presented in the emissions table was taken from [EIA's Monthly Report](https://www.eia.gov/totalenergy/data/monthly/#environment). 

```{r load-packages, message=FALSE, echo=FALSE}
#The analysis that follows was performed using R with the following libraries:
library(knitr)
library(plyr)
library(dplyr)
library(ggplot2)
library(reshape2)
```

```{r set-fig-path, echo=FALSE}
df <- read.csv('US_Electricity_Production_Sources_Percentages.csv')

df[,1] <- as.character(df[[1]])
for (i in 2:13) {
  df[,i] <- as.numeric(sub("%", "",df[[i]],fixed=TRUE)) / 100
}

opts_knit$set(base.dir = getwd(), base.url = '/')
opts_chunk$set(fig.path = 'assets/electricity-sources/')
```

## Dataset

The net generation dataset contains the electricity production in the United States by source.

The sources included are classified as: Coal, Petroleum Liquids, Petroleum Coke, Natural Gas, Other Gas, Nuclear Hydroelectric Conventional, Solar, Other Renewables, Hydroelectric Pumped, Storage, Other.

From the website: 

> Renewable Sources include wood, black liquor, other wood waste, biogenic municipal solid waste, landfill gas, sludge waste, agriculture byproducts, other biomass, geothermal, solar thermal, photovoltaic energy, and wind.

> Other includes non-biogenic municipal solid waste, batteries, hydrogen, purchased steam, sulfur, tire-derived fuel, and other miscellaneous energy sources.

## Total Generation

The generation of electricity has been mostly flat in the past ten years except for a small dip in 2009 due to the Great Recession. This is interesting because the dip in lost output (GDP) during the great recession was a lot deeper than the one shown on this graph.

```{r echo=FALSE}
ggplot(df, aes(Period, Total.Generation.at.Utility.Scale.Facilities, group = 1)) +
  geom_line(stat="identity") +
  expand_limits(y = 20000) +
  ggtitle(label = "Production and Direct Use", subtitle = NULL) +
  labs(x = "Period", y = "Quantity")
```

## Generation by Source

The following graph shows how the generation by source changed over time. Here are the trends over a ten year period ending in 2016:

- Coal has decreased from about 50% to about 30%.
- Natural Gas has increased from about 20% to above 30%
- Other Renewables increased from about 2% to about 5%
- Solar went from almost zero (0.01%) to almost 1%
- Nuclear has been constant around 20%

```{r echo=FALSE}
df <- plyr::rename(df, c(
  "Renewable.Sources.Excluding.Hydroelectric.and.Solar"="Other Renewables"))
mf <- melt(df[1:12], id.vars = 'Period', variable.name = 'Source', value.name = 'Quantity')


ggplot(mf, aes(Period, Quantity)) +
  geom_bar(aes(fill = Source), stat="identity") +
  ggtitle(label = "Electricity by Source", subtitle = NULL) +
  labs(x = "Year", y = 'Quantity (Percentage)')

```

## Coal vs Natural Gas

The following graph shows how coal is being replaced by natural gas and other renewables.

```{r echo=FALSE}
bp <- ggplot(df, aes(x=Period, group = 1)) +
  geom_line(aes(y = Coal, color = "Coal")) +
  geom_line(aes(y = Natural.Gas, color = "Natural Gas")) +
  geom_line(aes(y = df$`Other Renewables`, color = "Other renewables")) +
  labs(x = "Year", y = 'Quantity (Percentage)')
bp
```

## Carbon Dioxide Emissions by the Electricity Sector

In our period of interest the amount of carbon dioxide emitted by the electricity sector has been decreasing because of natural gas (which emits about 50% less than coal), and renewables (which does not emit any carbon dioxide).

```{r load-emissions, echo=FALSE}
# File was manually edited from https://www.eia.gov/totalenergy/data/browser/xls.php?tbl=T12.06&freq=m, which
# is part of the montly review at https://www.eia.gov/totalenergy/data/monthly/#environment
emissions <- read.csv('US_Emissions_Electricity_Sector.csv')
emissions <- filter(emissions, Period >= 2006)
```

```{r echo=FALSE, message=FALSE}
bp <- ggplot(emissions, aes(x=Period, Total.CO2.Emissions..MMT.)) +
  geom_line() +
  expand_limits(y = 0) +
  labs(x = "Year", y = 'Emissions (Millions of Metric Tons)')
bp
```
