---
title: 'Formula 1 Qualifying Battles'
output:
  md_document:
    pandoc_args: ["-s", "-t", "markdown"]
layout: post
---

## Introduction

This post is about evaluating the qualifying battles between Lewis Hamilton and Nico Rosberg, drivers who dominated the seasons from 2014 to 2016.

## Method

The most commonly used statistics when evaluating head to head qualifying battles is the simple count of how many times a driver outperformed the other. While this gives a pretty good idea of who was faster, more information is hidden in the gaps and outliers such as engine failures.

We will be looking at bar charts showing the race by race gaps between two drivers and characterize the gap distribution.

in order to evaluate the gaps fairly we have to remove outliers. We consider outliers circumstances when one of the two drivers does not make into Q3. It is safe to assume that a driver fighting for the championship will be in Q3 unless something really unexpected happens. We also remove gaps above two seconds because they are usually due to an interrupted lap or a mechanical failure.

```{r load-packages, message=FALSE, echo=FALSE}
#The analysis that follows was performed using R with the following libraries:
library(knitr)
library(dplyr)
library(ggplot2)
library(scales)
library(rjson)
library(purrr)
```

```{r set-fig-path, echo=FALSE}
opts_knit$set(base.dir = getwd(), base.url = '/')
opts_chunk$set(fig.path = 'assets/f1-quali/')
```

```{r read-date, echo=FALSE, warning=FALSE}
json_file <- function(driver, year) {
  paste('data/', driver, '_', year, '.json', sep = '')
}

or_na <- function(input) {
  if(!is.null(input) && input != "") input else NA
}

quali_data_frame <- function(driver, year) {
  json_string <- readLines(json_file(driver, year))
  json <- fromJSON(json_string)
  l <- lapply(json$MRData$RaceTable$Races, function(x) { 
    q <- x$QualifyingResults[[1]]
    c(x$Circuit$circuitId,
      x$Circuit$Location$country,
      as.numeric(q$position),
      or_na(q$Q1),
      or_na(q$Q2),
      or_na(q$Q3))
  })
  df <- setNames(do.call(rbind.data.frame, l), c("Circuit", "Race", "Position", "Q1", "Q2", "Q3"))
  df[, 3] <- as.numeric(as.character(df[, 3]))
  df
}
```

```{r, echo=FALSE, warning=FALSE}
to_seconds <- function(string) {
  if (!is.na(string)) {
    by_colon <- strsplit(string, split = ":")[[1]]
    min <- as.integer(by_colon[[1]])
    by_period <- strsplit(by_colon[[2]], split = "\\.")[[1]]
    sec <- as.integer(by_period[[1]])
    mil <- as.integer(by_period[[2]])
    60 * min + sec + mil / 1000
  } else {
    string
  }
}
```


```{r, echo=FALSE, warning=FALSE}
q3_times <- function(df) {
  times <- as.character(df$Q3) %>% map(to_seconds) %>% as.numeric()
  setNames(times, df$Race)
}

quali_times <- function(df) {
  times <- as.character(df$Q3) %>% map(to_seconds) %>% as.numeric()
  setNames(times, df$Race)
}

gaps <- function(df1, df2) {
  q3_1 <- q3_times(df1)
  q3_2 <- q3_times(df2)
  gaps <- q3_1 - q3_2 # TODO test case when races are not aligned
  removed <- gaps[is.na(gaps) | gaps >= 2]
  gaps <- gaps[!is.na(gaps)] # remove NAs
  gaps <- gaps[abs(gaps) < 2] # remove outliers (more than a two second gap)
  list(gaps = gaps, removed = removed)
}

compare_runs <- function(driver1, driver2, year) {
   q1 <- quali_data_frame(driver1, year)
   q2 <- quali_data_frame(driver2, year)
   g <- gaps(q1, q2)
   d1 <- sum(q1$Position < q2$Position)
   d2 <- sum(q1$Position > q2$Position)
   d1_wins <- sum(g$gaps < 0)
   d2_wins <- sum(g$gaps > 0)
   head_to_head <- paste(d1, ' to ', d2, sep = '')
   head_to_head_without_outliers <- paste(d1_wins, ' to ', d2_wins, sep = '')
   list(
     driver1 = driver1,
     driver2 = driver2,
     gaps = g$gaps,
     mean = mean(g$gaps),
     stddev = sd(g$gaps),
     head_to_head = head_to_head,
     head_to_head_without_outliers = head_to_head_without_outliers,
     removed = names(g$removed))
}

#compare_runs('alonso', 'button', 2016)


```

```{r, echo=FALSE, warning=FALSE}
gap_graph <- function(df) {
  y <- df$gaps
  x <- names(unlist(df$gaps))
  gap_df <- data.frame(x, y)
  gap_df$x <- factor(gap_df$x, levels = unique(gap_df$x))
  ggplot(gap_df, aes(x, y)) + geom_bar(stat = 'identity') +
    labs(x = "Race", y = "Gap") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
}
```

# Hamilton vs Rosberg (2016)

```{r, echo=FALSE, warning=FALSE}
ham_vs_ros_2016 <- compare_runs('hamilton', 'rosberg', 2016)
```

```{r, echo=FALSE, warning=FALSE}
gap_graph(ham_vs_ros_2016)
```

## Head to Head

The overall head to head saw Hamilton edging out Rosberg: ```r ham_vs_ros_2016$head_to_head```. After removing outliers, the head to head turns more into Hamilton's favor: ```r ham_vs_ros_2016$head_to_head_without_outliers```

## Outliers

Omitted from gaps calculations: ```r ham_vs_ros_2016$removed```

* China: A hybrid system failure in the turbo charger prevented Hamilton from setting a time in qualifying
* Russia: The same mechanical failure as in China prevented Hamilton from competing in Q3
* Azerbaijan: Hamilton crashed in Q3 before setting a quick lap time. While this is Hamilton's fault and deserves
to be counted as a Rosberg win, it doesn't make sense to use the gap and we consider it an outlier.
* Belgium: Hamilton took a penalty for changing the engine and did not participate in Q2 and Q3.

Three of the outliers are mechanical failures on Hamilton's car, while the remaining one is a very unusual crash in Q3 from Hamilton.

## Gap Distribution

Average gap: ```r ham_vs_ros_2016$mean```
Standard deviation: ```r ham_vs_ros_2016$stddev```

Hamilton was on average a bit over a tenth faster than Rosberg. When Rosberg was faster, he was usually by a very small margin except for Singapore, where we saw the biggest margin between the two drivers (over 0.7s).

# Hamilton vs Rosberg (2015)

```{r, echo=FALSE, warning=FALSE}
ham_vs_ros_2015 <- compare_runs('hamilton', 'rosberg', 2015)
gap_graph(ham_vs_ros_2015)
```

## Head to Head

The overall head to head saw Hamilton edging out Rosberg: ```r ham_vs_ros_2015$head_to_head```. After removing outliers, the gap between the two drivers is wider: ```r ham_vs_ros_2015$head_to_head_without_outliers```.

This result ius very similar to the 2016 season. Also to be noted that Rosberg only outqualified Hamilton once in the first thirteen races, and then outqualified Hamilton every single time to the end of the season.

## Outliers

Omitted from gaps calculations: ```r ham_vs_ros_2015$removed```

-- TODO use Q2 results

## Gap Distribution

Average gap: ```r ham_vs_ros_2015$mean```
Standard deviation: ```r ham_vs_ros_2015$stddev```

Hamilton was on average a bit over a tenth faster than Rosberg, a result very close to 2016.

# Hamilton vs Rosberg (2014)

```{r, echo=FALSE, warning=FALSE}
ham_vs_ros_2014 <- compare_runs('hamilton', 'rosberg', 2014)
gap_graph(ham_vs_ros_2014)
```

## Head to Head

The overall head to head this time saw Hamilton being edged out by Rosberg: ```r ham_vs_ros_2014$head_to_head```. After removing outliers, the result does not change but it is definitely closer: ```r ham_vs_ros_2014$head_to_head_without_outliers```.

## Outliers

Omitted from gaps calculations: ```r ham_vs_ros_2014$removed```

* Austria: Hamilton spun on his second Q3 lap after his first lap was deleted because he exceeded the track limits
* United Kingdom: Q3 was a wet session. Hamilton set his time too early and the gap between Rosberg and Hamilton was over two seconds
* Germany: Hamilton crashed in Q1 due to a brake disc failure
* Hungary: Hamilton's car caught fire during qualifying

## Gap Distribution

Average gap: ```r ham_vs_ros_2014$mean```
Standard deviation: ```r ham_vs_ros_2014$stddev```

Hamilton was on average a bit over a tenth faster than Rosberg, a result very close to 2016.

# All Together

<div markdown="0">
```{r echo=FALSE}
battles <- list(
  b2016 = ham_vs_ros_2016,
  b2015 = ham_vs_ros_2015,
  b2014 = ham_vs_ros_2014)

df <- data.frame(
  year = c("2016", "2015", "2014"),
  drivers = unlist(map(battles, function(x) x$driver1)),
  avg = unlist(map(battles, function(x) x$mean)),
  stddev = unlist(map(battles, function(x) x$stddev)))
kable(df, 
      col.names = c("Year", "Drivers", "Average Gap", "Std Dev"),
      format = 'html',
      table.attr = "class=\"table\"")
```
</div>